Interpretable Neural Networks using EAGGA

### 2025-03-15 paper/main.tex
conclusion

%likely shouldn't be an entire section
\section{EAGGA}
EAGGA, as proposed by \citet{EAGGA}, is an NSGA-II inspired algorithm that optimizes performance along with three interpretability objectives.
The three interpretability proxies are
\begin{itemize}
  \item NF: the relative number of features used in the model
  \item NI: the relative number of pairwise interaction effects in the model
  \item NNM: the relative number of non-monotone feature effects in the model
\end{itemize}
The authors argue that a sparse model as indicated by a low NF score helps interpretability by reducing the number of e.g. ALE plots necessary in post-hoc
analysis \citep[p. 540]{EAGGA}. Similarly, a low NI score reduces the interplay between features and thus the dimensionality of post-hoc analysis plots
and a low NNM avoids complex non-monotone feature effects, which is often desirable in practical use-cases such as loan approval, where a higher income should
translate to a higher chance of approval.
Lastly, performance is measured by the area under the receiver operating characteristic curve (AUC), which trades-off the true against the false positives rate.
\\
\citet[pp. 540f]{EAGGA} acknowledge that this set-up creates the complex ``extended search space'' $\check{\boldsymbol\Lambda}$ to be optimized over, containing
tuples $(\boldsymbol\lambda, \boldsymbol{s}, \boldsymbol{I}_s, \boldsymbol{m}_{\boldsymbol{I}_{\boldsymbol{s}}})=\check{\boldsymbol\lambda}\in\check{\boldsymbol\Lambda}$
with
\begin{itemize}
  \item $\boldsymbol\lambda\in\boldsymbol\Lambda$ being the model's hyperparameter configuration
  \item $\boldsymbol{s}\in\{0,1\}^p$ being a binary vector denoting which features are used in the model
  \item $\boldsymbol{I}_s\in\{0,1\}^{p\times p}$ being a binary matrix denoting included pairwise interactions between features
  \item $\boldsymbol{m}_{\boldsymbol{I}_{\boldsymbol{s}}}\in\{-1,0,1\}^p$ being the vector denoting the monotonicity constraint of a feature (-1 decreasing, 0 none, 1 increasing)
\end{itemize}
Consequently, the optimization problem over this space would be
\begin{equation}
  \argmin_{\check{\boldsymbol\lambda}\in\check{\boldsymbol\Lambda}} \big(
    GE(\mathcal{I}_{\check{\boldsymbol\lambda}},\mathcal{D}),
    NF(\hat{f}_{\mathcal{D},\check{\boldsymbol\lambda}}),
    NI(\hat{f}_{\mathcal{D},\check{\boldsymbol\lambda}}),
    NNM(\hat{f}_{\mathcal{D},\check{\boldsymbol\lambda}})
  \big)
  \label{eq-eagga-extended-space}
\end{equation}
The authors thus introduce the equivalence relation $R$ ``allowed to interact'' on the set of features $C$. They further denote all features included in the model with $C_s$.
This induces a group structure of equivalence classes, where features allowed to interact belong to the same class and share the same monotonicity attribute.
Using this group structure, they define a much simpler ``augmented serach space'' $\tilde{\boldsymbol\Lambda}=\boldsymbol\Lambda\times\mathcal{G}$ comprising of
the cartesian product of the model hyperparameter space $\boldsymbol\Lambda$ and the group structure space $\mathcal{G}$.
Each group structure $\boldsymbol{G}\in\mathcal{G}$ consists of $g$ groups:
\begin{itemize}
  \item $G_1=C\setminus C_s$ being the set of excluded features
  \item $G_k=(E_k,M_{E_k}), \forall k=2,...,g$ being tuples with
  \begin{itemize}
    \item $E_k\subseteq C_s$ being the set of features in group $k$, i.e. features allowed to interact with each other
    \item $M_{E_k}$ being the monotonicity constraint of group $k$
  \end{itemize}
\end{itemize}
To generate new offspring, they use a regular evolutionary algorithm on the model hyperparameters $\boldsymbol\Lambda$ and a grouping genetic algorithm (GGA) on
the group structure space $\mathcal{G}$, with adapted mutation and crossover operators.
The authors further introduce a special initialisation of the group structures to increase the sample-efficiency of their algorithm compared to random initialisation.
\citep[pp. 541-543]{EAGGA}

@@

neighbouring fronts, which seemed reasonable not to be concerned over.  % ok formulation?

@@

\section{Conclusion and Future Outlook}
(Conclusion)
- tabular data still difficult discipline for neural networks
- research suggests heavy regularisation to improve performance on tabular data
- EAGGA proved to be successful in making XGBoost more interpretable while keeping performance on par with unregularised XGBoost
- we extend EAGGA to neural networks to see if we can utilise the regularisation induced by it to make NNs both interpretable and performant on tabular data
- as consequence propose new architecture allowing to model equivalence relations of EAGGA
- ... conforming to \citet[chap. 2]{survey_NN_interpretability} taxonomy it fits as ..., ..., ... %TODO
- found overall performance comparable to that of XGBoost fitted using EAGGA, which is a plus, but no outperformance

(Future Outlook)
- MO BO on group structure space possible, perhaps via restricted BO?

@@

%\newpage  % we are not to add page breaks

>>>>

%likely shouldn't be an entire section
\section{EAGGA}
\label{sec-eagga}
EAGGA, as proposed by \citet{EAGGA}, is an NSGA-II inspired algorithm that optimizes performance along with three interpretability objectives.
The three interpretability proxies are
  \item NF: the relative number of features used in the model
  \item NI: the relative number of pairwise interaction effects in the model
  \item NNM: the relative number of non-monotone feature effects in the model
The authors argue that a sparse model as indicated by a low NF score helps interpretability by reducing the number of e.g. ALE plots necessary in post-hoc
analysis \citep[p. 540]{EAGGA}. Similarly, a low NI score reduces the interplay between features and thus the dimensionality of post-hoc analysis plots
and a low NNM avoids complex non-monotone feature effects, which is often desirable in practical use-cases such as loan approval, where a higher income should
translate to a higher chance of approval.
Lastly, performance is measured by the area under the receiver operating characteristic curve (AUC), which trades-off the true against the false positives rate.
\citet[pp. 540f]{EAGGA} acknowledge that this set-up creates the complex ``extended search space'' $\check{\boldsymbol\Lambda}$ to be optimized over, containing
tuples $(\boldsymbol\lambda, \boldsymbol{s}, \boldsymbol{I}_s, \boldsymbol{m}_{\boldsymbol{I}_{\boldsymbol{s}}})=\check{\boldsymbol\lambda}\in\check{\boldsymbol\Lambda}$
with
  \item $\boldsymbol\lambda\in\boldsymbol\Lambda$ being the model's hyperparameter configuration
  \item $\boldsymbol{s}\in\{0,1\}^p$ being a binary vector denoting which features are used in the model
  \item $\boldsymbol{I}_s\in\{0,1\}^{p\times p}$ being a binary matrix denoting included pairwise interactions between features
  \item $\boldsymbol{m}_{\boldsymbol{I}_{\boldsymbol{s}}}\in\{-1,0,1\}^p$ being the vector denoting the monotonicity constraint of a feature (-1 decreasing, 0 none, 1 increasing)
Consequently, the optimization problem over this space would be
  \argmin_{\check{\boldsymbol\lambda}\in\check{\boldsymbol\Lambda}} \big(
    GE(\mathcal{I}_{\check{\boldsymbol\lambda}},\mathcal{D}),
    NF(\hat{f}_{\mathcal{D},\check{\boldsymbol\lambda}}),
    NI(\hat{f}_{\mathcal{D},\check{\boldsymbol\lambda}}),
    NNM(\hat{f}_{\mathcal{D},\check{\boldsymbol\lambda}})
  \big)
  \label{eq-eagga-extended-space}
The authors thus introduce the equivalence relation $R$ ``allowed to interact'' on the set of features $C$. They further denote all features included in the model with $C_s$.
This induces a group structure of equivalence classes, where features allowed to interact belong to the same class and share the same monotonicity attribute.
Using this group structure, they define a much simpler ``augmented serach space'' $\tilde{\boldsymbol\Lambda}=\boldsymbol\Lambda\times\mathcal{G}$ comprising of
the cartesian product of the model hyperparameter space $\boldsymbol\Lambda$ and the group structure space $\mathcal{G}$.
Each group structure $\boldsymbol{G}\in\mathcal{G}$ consists of $g$ groups:
  \item $G_1=C\setminus C_s$ being the set of excluded features
  \item $G_k=(E_k,M_{E_k}), \forall k=2,...,g$ being tuples with
    \item $E_k\subseteq C_s$ being the set of features in group $k$, i.e. features allowed to interact with each other
    \item $M_{E_k}$ being the monotonicity constraint of group $k$
To generate new offspring, they use a regular evolutionary algorithm on the model hyperparameters $\boldsymbol\Lambda$ and a grouping genetic algorithm (GGA) on
the group structure space $\mathcal{G}$, with adapted mutation and crossover operators.
The authors further introduce a special initialisation of the group structures to increase the sample-efficiency of their algorithm compared to random initialisation.
\citep[pp. 541-543]{EAGGA}

@@

neighbouring fronts, hence we considered it tolerable.

@@

\section{Conclusion and Future Outlook}
% Conclusion
The EAGGA framework has been shown to be able to preserve performance while increasing interpretabiltiy of XGBoost models on tabular data, which has long been 
a challenging discipline for neural networks. Following recent research suggesting that regularisation can improve
performance \citep[8]{NEURIPS2021_c902b497}, we extended EAGGA \citep{EAGGA} to neural networks, utilising its regularising properties to achieve similar
performance of neural networks as performance achieved with XGBoost and simultaneously increase interpretability of this notoriously uninterpretable model class.
As a consequence, we propose a new architecture allowing to model the equivalence relations of the EAGGA framework.
We create a model whose interpretation is active, attributional, and global in the framework of \citet[chap. 2]{survey_NN_interpretability}, and find
overall performance, as measured on the dominated hypervolume of (AUC, NF, NI, NNM) to be comparable to that of XGBoost fitted usign EAGGA.

(Future Outlook)
- due to high complexity of EAs, a MOBO implementation on augmented space (cf section \ref{sec-eagga}) would be very nice
- theoretical idea: use surrogate model with prior computed based on feasible combinations of equivalence classes
  * high value, 0 uncertainty around combinations that make no sense
    - e.g. those with a feature in multiple groups
    - e.g. those with two features in one group, one having monot incr, one monot decr
  * high uncertainty around combinations that would make sense to encourage exploring these regions as with GPs
  * this would likely prove to be computationally very expensive for higher numbers of features
- as more practical alternative: unrestricted MO BO on extended search space, cf section \ref{sec-eagga}, then acquisiton function taking into
  account the constraints, e.g. as presented by \citep{10.5555/3020751.3020778}, \citep{perrone2019constrainedbayesianoptimizationmaxvalue}

@@
